---
title: "Lab 04 - That all you got?  Pretty much mastered this stuff-ola"
output: html_document
---

### Using real data -- weeeeeeeeeeee! Fun! 
R includes a number of datasets that it is convenient to use for examples. You can get a description of what's available by typing
```{r}
data()
```
To access any of these datasets, you then type data(dataset) where dataset is the name of the
dataset you wish to access. For example,
```{r}
data(trees)
```
Typing
```{r}
trees[1:5,]
```
gives us the first 5 rows of these data, and we can now see that the columns represent measurements
of girth, height and volume of trees (actually cherry trees: see help(trees)) respectively. Now, if we want to work on the columns of these data, we can use the subscripting technique
explained above: for example, trees[,2] gives all of the heights. 

What is the average height of the trees in the trees dataset?
```{r, echo=FALSE}
data(trees)
mean(trees$Height)
```


The $ allows you extract elements by name from a named list. For example
```{r}
x <- list(a=1, b=2, c=3)
x$b
```
You can find the names of a list using names()
```{r}
names(x)
```
This is a basic extraction operator. You can view the corresponding help page by typing ?Extract in R.
```{marginfigure}
PLEASE: make sure we talk about the risk/disadvantage of attaching data!
```

This is a bit tedious however, and it would be easier if we could refer to the heights more explicitly. We can achieve this by attaching to the trees dataset:
```{r}
attach(trees)
```
Effectively, this makes the contents of trees a directory, and if we type the name of an object, R
will look inside this directory to find it. Since Height is the name of one of the columns of trees, R now recognizes this object when we type the name. Hence, for example,
```{r}
mean(Height)
mean(trees[,2])
```
are synonymous, while it is easier to remember exactly what calculation is being performed by
the rst of these expressions. In actual fact, trees is an object called a dataframe, essentially a
matrix with named columns (though a dataframe, unlike a matrix, may also include non-numerical
variables, such as character names). Because of this, there is another equivalent syntax to extract, for example, the vector of heights:
```{r}
trees$Height
```
which can also be used without having first attached to the dataset.

### Exercises
1. Attach to the dataset quakes and produce a statistical summary of the variables depth and
mag.
2. Attach to the dataset mtcars and find the mean weight and mean fuel consumption for
vehicles in the dataset (type help(mtcars) for a description of the variables available).
