---
title: "Lab 05 - I can't think of anything more rewarding than being able to express yourself to others through painting. - Bob Ross"
output: html_document
---


```{marginfigure}
Later in the course (maybe next course) we will use a really nice tool for graphics: ggplot2. ggplot2 is a system for declaratively creating graphics, based on The Grammar of Graphics. You provide the data, tell ggplot2 how to map variables to aesthetics, what graphical primitives to use, and it takes care of the details. For now though, just a few ou-of-the-box basics. 
```

### Graphics
R has many facilities for producing high quality graphics. 

A useful facility before beginning is to divide a page into smaller pieces so that more than one gure can be displayed. For example:
```{r}
par(mfrow=c(2,2))
```

creates a window of graphics with 2 rows and 2 columns. With this choice the windows are filled
up row-wise. Use mfcol instead of mfrow to fill up column-wise. The function par is a general
function for setting graphical parameters. There are many options: see help(par).
```{r}
par(mfrow=c(2,2))
attach(trees)
hist(Height)
boxplot(Height)
hist(Volume)
boxplot(Volume)
```
Note the final use of par to return the graphics window to standard size.
```{r}
par(mfrow=c(1,1))
```

We can also plot one variable against another using the function plot:
```{r}
plot(Height,Volume)
```
R can also produce a scatterplot matrix (a matrix of scatterplots for each pair of variables)
using the function pairs:
```{r}
pairs(trees)
```
Like many other functions plot is object-specific: its behavior depends on the
object to which it is applied. For example, if the object is a matrix, plot is identical to pairs: try
plot(trees). For some other possibilities try:

1. data(nhtemp)
2. plot(nhtemp)
3. data(faithful)
4. plot(faithful)
5. data(HairEyeColor)
6. plot(HairEyeColor)

There are also many optional arguments in most plotting functions that can be used to control
colors, plotting characters, axis labels, titles etc. The functions points and lines are useful for
adding points and lines respectively to a current graph. The function abline is useful for adding
a line with specified intercept and slope.

To print a graph, point the cursor over the graphics window and press the right button on the
mouse. This should open up a menu which includes `print' as an option. You also have the option
to save the figure in various formats, for example as a postscript file, for storage and later use.

### Exercises
1. Use
  > x<-rnorm(100)
or something similar, to generate some data. Produce a figure showing a histogram and
boxplot of the data. Modify the axis names and title of the plot in an appropriate way.
2. Type the following -- what awesomeness do you get?
```{r, eval=FALSE}
x<- (-10):10
n<-length(x)
y<-rnorm(n,x,4)
plot(x,y)
abline(0,1)
```

Try to understand the effect of each command and the graph that is produced.
3. Type the following:
```{r, eval=FALSE}
data(nhtemp)
plot(nhtemp)
```
This produces a time series plot of measurements of annual mean temperatures in New Hampshire, U.S.A.
4. The previous example illustrated that plot acts dierently on ob jects of dierent types - the
object nhtemp has the special class time series. More generally, we may have the data of
yearly observations in a vector, but need to build the time series plot for ourselves. We can
mimic this situation by typing
```{r, eval=FALSE}
temp<-as.vector(nhtemp))
```
which creates a vector temp that contains only the annual temperatures. We can produce
something similar to the time series plot by typing
```{r, eval=FALSE}
plot(1912:1971,temp)
```
but this plots points rather than lines. To join the data via lines we would use
```{r, eval=FALSE}
plot(1912:1971,temp)
plot(1912:1971,temp,type='l')
```
instead. To get points and lines, use type='b' 
